What Drives Emergency Medical Services (EMS) Demand?

### 2024-09-27 paper/paper.qmd
Reviewed and edited entire paper again.

In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" [@auditorreport] of Toronto's paramedic services, further research on how to manage EMS demand is needed. Thus, by using Paramedic Services Incident Data offered by Toronto Paramedic Services, this analysis seeks to identify which types of emergencies (e.g. medical emergencies or vehicle accidents) drive EMS demand. Knowing the drivers of EMS demand can help inform paramedic service strategies that allow for better EMS demand management.

@@

Paramedic Services Incident Data (2017-2022) is provided by Toronto Paramedic Services [@torontoparamedicservices]. The records of when a paramedic was dispatched having received a call, they type of incident, the priority level of each incident, the number of paramedic units that arrived at the scene, and the general location of each incident are all updated by hand to the city of Toronto's Open Data portal [@opentorontodata].

@@

The Paramedic Services Incident Data is refreshed annually and contains information on every single incident paramedics have responded to since January 1st to December 31st in 2017 to 2022. For every day of each year, there are multiple records of incidents that have occurred. For every record, there is an ID number, the date and time when a paramedic unit was dispatched, the type of incident, the severity of the incident (priority), the number of paramedic units at the scene, and the general location of the incident. @tbl-rawdata below previews this information for recorded incidents in 2022.

@@

To further unpack what is driving this consistent demand, the number of emergency calls for individual incident type are also considered. @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls. Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are key drivers that propel EMS demand.

@@

By zooming into @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided. With @fig-incidentemsdemandzoom, it becomes clear that vehicle accidents and emergency transfers may be other potential drivers of EMS demand, though not as strong as medical emergencies.

@@

## Mapping EMS Demand

@@

These results support the findings from @sec-incidenttypes. Medical emergencies, emergency transfers, and vehicle accidents are shown to be relevant drivers of EMS demand. The findings further imply that what drives EMS demand is dependent on the characteristics of a particular region. Socio-economic or demographic factors such as income or population density of a given area can impact the types of incidents that occur, total number of emergency calls that are made and the severity of each incident. Given this, general strategies that target the three main drivers of EMS demand may not be effective for all communities. Thus, there is a need to consider the characteristics of different communities to build community-based strategies that effectively address the main drivers of EMS demand.

@@

To further this research, the relationship between the driver of EMS demand and socio-economic or demographic characteristics for each forward sortation area should be examined. As there are many variables that can impact one's health and ultimately, whether or not they have to seek EMS, understanding how factors like income or population density affect the occurrence of medical emergencies, emergency transfers, and vehicle accidents can allow for the formation of effective community-based strategies. It may also be helpful to consider ambulance station locations in addition to @fig-map to see whether ambulance stations and resources are optimally distributed and if additional resources are needed depending on the EMS demand driver for a particular community.

@@

Though this analysis carefully examines EMS demand, there are some limitations to consider. It is mentioned in @sec-analysis that the records of incidents without an assigned incident type were omitted from the analysis. This may have limited the overall representation of the number of emergency calls for each incident type. Further, it is important to note that @fig-map only considers data from 2022. So, although @fig-map demonstrates that differences between communities and their drivers for EMS demand can exist, a better reflection of the actual drivers of EMS demand for each community may be achieved by considering multiple years of incident data (i.e. from 2017 to 2022).

@@

To isolate the forward sortation areas given in the Paramedic Services Incident Data for 2022, ArcGIS Pro software [@arcgis] was used to remove any areas that were not recorded. Specifically, the forward sortation areas that were not present within the Paramedic Services Incident Data were deleted from the shapefile's attribute table (i.e. data table) and re-saved as a new shapefile. After making the necessary modifications, the spatial data for forward sortation areas was merged with the data present in @tbl-mapdata and mapped to show the "EMS Demand Driver" for every area. The final map is shown in @fig-map.

>>>>

In response to this rising challenge, the report posits that in order to elevate the "efficiency and effectiveness" [@auditorreport] of Toronto's paramedic services, further research on how to manage EMS demand is needed. Thus, by using Paramedic Services Incident Data offered by Toronto Paramedic Services, this analysis seeks to identify which types of emergencies (e.g. medical emergencies, vehicle accidents, etc.) drive EMS demand. Knowing the drivers of EMS demand can help inform paramedic service strategies that allow for better EMS demand management.

@@

Paramedic Services Incident Data (2017-2022) is provided by Toronto Paramedic Services [@torontoparamedicservices]. The records of when a paramedic was dispatched having received a call, they type of incident, the priority level of each incident, the number of paramedic units that arrived at the scene, and the general location of each incident are all periodically updated by hand to the city of Toronto's Open Data portal [@opentorontodata].

@@

The Paramedic Services Incident Data is refreshed annually and contains information on every single incident paramedics have responded to since January 1, 2017 to December 31, 2022. For every day of each year, there are multiple records of incidents that have occurred. For every record, there is an ID number, the date and time when a paramedic unit was dispatched, the type of incident, the severity of the incident (priority), the number of paramedic units at the scene, and the general location of the incident. @tbl-rawdata below previews this information for recorded incidents in 2022.

@@

To further unpack what is driving this EMS demand, the number of emergency calls for individual incident type are also considered. @fig-incidentemsdemand illustrates the number of emergency calls for each incident type along with the total number of emergency calls. Based on @fig-incidentemsdemand, demands for EMS are markedly higher for medical emergencies than any other incident type, suggesting that medical circumstances are key drivers that propel EMS demand.

@@

By zooming into the lower portion of @fig-incidentemsdemand with @fig-incidentemsdemandzoom below, a clearer look at the other incident types is provided. With @fig-incidentemsdemandzoom, it becomes clear that vehicle accidents and emergency transfers may be other potential drivers of EMS demand, though not as strong as medical emergencies.

@@

## Mapping EMS Demand {#sec-maplocation}

@@

These results support the findings from @sec-incidenttypes. Medical emergencies, emergency transfers, and vehicle accidents are shown to be relevant drivers of EMS demand. The findings further imply that what drives EMS demand is dependent on the characteristics of a particular region. Socio-economic or demographic factors such as income or population density of a given area can impact the types of incidents that occur, total number of emergency calls that are made and the severity of each incident. Given this, general, blanketing strategies that target the three main drivers of EMS demand may not be effective for all communities. Thus, there is a need to consider the characteristics of different communities to build community-based strategies that effectively address the main drivers of EMS demand.

@@

To further this research, the relationship between the driver of EMS demand and socio-economic or demographic characteristics for each forward sortation area should be examined. As there are many variables that can impact one's health and ultimately, whether or not they have to seek EMS, understanding how factors like income or population density affect the occurrence of medical emergencies, emergency transfers, and vehicle accidents can allow for the formation of effective community-based strategies. It may also be helpful to consider ambulance station locations in addition to @fig-map in @sec-maplocation to see whether ambulance stations and resources are optimally distributed and if additional resources are needed based on the EMS demand driver for a particular community.

@@

Though this analysis carefully examines EMS demand, there are some limitations to consider. It is mentioned in @sec-analysis that the records of incidents without an assigned incident type were omitted from the analysis. This may have limited the overall representation of the number of emergency calls for each incident type. Further, it is important to note that @fig-map only considers data from 2022. So, although @fig-map in @sec-maplocation demonstrates that differences between communities and their drivers for EMS demand can exist, a better reflection of the actual drivers of EMS demand for each community may be achieved by considering multiple years of incident data (i.e. from 2017 to 2022).

@@

To isolate the forward sortation areas given in the Paramedic Services Incident Data for 2022, ArcGIS Pro software [@arcgis] was used to remove any areas that were not recorded. Specifically, the forward sortation areas that were not present within the Paramedic Services Incident Data were deleted from the shapefile's attribute table (i.e. data table) and re-saved as a new shapefile. After making the necessary modifications, the spatial data for forward sortation areas was merged with the data present in @tbl-mapdata and mapped to show the "EMS Demand Driver" for every area. The final map is shown in @fig-map in @sec-maplocation.
